Network based multifactorial modelling microRNA-target interaction.

### 2019-04-07 article.Rmd
cross-ref

MicroRNAs (miRNAs) that are key regulator of gene expression through various post-transcriptional mechanisms are a family of short non-coding RNAs. Although repression mechanisms of miRNA are not fully understood, exact effect of miRNAs is regulation of direct target/targets negatively. Repressive activities of miRNAs vary depending on many factors that are significant to microRNA:target interactions. These factors include microRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energy of miRNA:target complexes varies based on nucleotide context and determines folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’UTR bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown miRNAs can interact with targets via sequence that located in different region such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013]moore_mirnatarget_2015. These studies also shown that binding location could indicate functionality of miRNA:target interaction or affect profile of target abundance. It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is prefectly complementary with targets, seed region at the 5’UTR of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides and locates in different region of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. In addition to all of these, abundance of miRNAs and targets and miRNA:target ratio in cells are determinant for efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. 
As it is possible for miRNAs to suppress multiple targets, an individual RNA molecule can also be targeted by various miRNAs. In that case, the targeted RNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbtion of the other target when a steady-state system that included one miRNA and two target, was perturbed with expression changes at one of the targets [@ala_integrated_2013]. When interaction between miRNAs and their target were dealed in the cell, explanation and prediction of results of perturbation is difficult due to varieties of interactions. Various computational and experimental studies have applied to be understood for ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other target expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping to affinity of targets. They have demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets also were studied and it has been demonstrated crucial for regulations at expression of targets [@denzler_impact_2016]. MiRNA:target interactions have modelled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended to handling models in network structures [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA-target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that whole miRNAs and targets in network can interacted through common miRNAs and genes [@robinson_modelling_2018. Associated genes that is targets of the same miRNAs have been found with help of correlation of gene expression changes in recent study [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA’s studies and published these approach as a package that can use with R programming.

@@

We developed a novel network based approach that can use to simulation of gene regulation after any perturbation at miRNA:target interactions. If a perturbation or change induces on graph, whole interacted nodes vary because of changes in miRNA:target ratios. The system was assumed as steady state that miRNAs affect according to the ratio of the individual target to all targets (Fig.1a, expression values shown as integer for comprehensible sampling). When expression of gene 2 (G2) increased to two fold (Fig. 1b), distribution of miRNA (M1) expression changes (Fig. 1c). Consequently, expression of common target (G4) of miRNA1 and miRNA2 also varies. The change at expression of common target triggers to changes of proportional distribution of miRNA2 (Fig. 1d).

![Shematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangle grey) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed. ](Fig1.jpg)

@@

$$Eff_{gi}= C_m * C_{gi}/\sum_{0}^{i} C_g $$
$$R_{gi}= Eff_{i1}+ Eff_{i2} ... $$

The efficiency of a miRNA on the individual target (Effgi) is calculated according to i; Cm, miRNA expression in the group; Cgi, symbolize individual gene expression; Cg gene expressions in the group (Groups are determined depending on miRNAs. Targets that are interacted with the same miRNA find in a group.). In this approach, because a gene can be found in different groups, the cooperative activity of microRNAs also can be calculated. Total repression activity of different miRNAs on the common target, R, is calculated based on ii; sum of efficiency of different miRNAs on the target.

![Calculations for determine to miRNA binding and repression efficiency. G, Gene; M, miRNA; STE, seed type effect; RE, Region Effect; E, Energy; STE’, normalized values of seed type efficiency coefficient; RE’, normalized values of region efficiency coefficient; E’, normalized values of energy coefficient.](fig2.jpg)

@@

Interactions between miRNAs and their targets can be affected from different parameters. For this reason, we improved model with factors that can affect to miRNA activity. In our approach, the parameters are classified as binding and efficiency factors and all of these are evaluated in their groups. Binding factors determine interaction between miRNA and target. The efficiency factors decide functionality of binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017]in miRNA:target interactions are represented as binding affinity. Efficiency factors determine how many of microRNA:target complexes will result in inhibition. For example, binding region on the target shows whether the target can be inhibited (functionality)[@hausser_analysis_2013; @helwak_mapping_2013]. Both of binding and efficiency factors are evaluated in the groups and normalized according to iii; each factor is normalized based on maximum variables that found in their groups.

$$F`= F/F_{max}$$

The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (Fig 2).  Binding affinities (activity, Eff) of miRNAs on each individual gene are calculated as shown in iv; Cm, miRNA expression in the group; Cg, Gene expression; Cgi, individual gene expression; gi, individual gene; g, whole genes in a group (Fig 2c).

$$Eff_{gi}= C_m * E`_{gi}*STE`_{gi}*C_{gi}/(\sum_{0}^{i} E`_{gi}*STE`_{gi}*C_{gi}) $$

$$Eff_{gi}= Eff_{gi}*RE`_{gi} $$

After miRNA binds to its target, but might not repress to bound target. The functionality of bound miRNA on target depends on efficiency factors like region that is binding sequence of miRNA on its target. Exact repression efficiency of miRNA is calculated according to equation v (Fig 2d); RE’gi, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Fig 2e.

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow.jpg)

@@

We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous sections. The functions and steps of approach are explained as following:
1-2. Convertion of dataset: priming_graph function process the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (i-v) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state. 
2-3. Transition of variables in graph: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.
3-4. Trigger change in graph: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The fist, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.
4-5. Updating current values of variables: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.
5-6. Simulation of competing behavior of targets: After the change in the steady- state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.
The node elements in the dataset are handled as two type; repressives (miRNAs) and competings (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the approach are developed with R programming so as can be used with other packages. These are can be found in the github repository [ceRNAnetsim github page](https://github.com/selcenari/ceRNAnetsim) and improved with contributions of others.

@@

We developed a model approach and workflow for competitive ceRNA regulation in this study. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. For this reason, we performed first application as considering only abundance of miRNAs and targets (Fig 1). After the increasing in a gene expression value, expression values of other genes also changed slightly in comparison with alteration in perturbation started gene node. Different studies have shown that the other gene expression values also rise differently, if a gene abundance increases in ceRNA system [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (Fig 1d). The common target in system displayed like a trigger for the other group and induced changes of expression values of genes on the other group. Therefore, effect of primary change of the gene (G2) on genes of the other group was observed slightly (Fig 1d). In addition, like shown by ceRNA hypothesis model of Ala et al., after the increasing of a gene expression (G2), the miRNA that is found in the same group (M1) tended to be more repressive on this target. This caused to decrease in increased expression value of primer triggering gene.
When the expression of a gene (G2) was increased to two fold, we observed that the expression changes were more evident. So, we considered that high miRNA:target ratio results in stronger miRNA repression activity. This results also are coherent with previous reports that were offered by Arvey et al. and Denzler et al.[@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. When the simulation was ran in sample dataset that includes lower target abundance, the first response of primary neihgborhood of the same group was determined as similar with in system that have high target abudance. But regulations of ceRNAs were observed differences in the folowing steps of simulation. In overall state in the end of simulation, we established that all regulations more prominent when compared with low miRNA:target ratio datset. The model appeared the importance of mRNA:target ratio in regulation of competing RNAs.

![Target regulations with interaction parameters. a) In the steady-state the repression activity of miRNAs on the targets after binding and repression efficiency. b) The changes the repression activities after increasing of G2 expression. c) Perturbation of primary neighborhoods of M1 miRNA (M1 miRNA group). d) Regulation of gene expression of other gene group via triggering target (common target between M1 and M2). ](Fig4.jpg)

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We explained the model in the same sample dataset with using different factors and simulated via the same change (Fig 4). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Fig 4a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (Fig 4 b,c,d).
In regulation of miRNA target sample system (Fig 4), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2.  When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions were observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation.

@@

>>>>

MicroRNAs (miRNAs) that are key regulator of gene expression through various post-transcriptional mechanisms are a family of short non-coding RNAs. Although repression mechanisms of miRNA are not fully understood, exact effect of miRNAs is regulation of direct target/targets negatively. Repressive activities of miRNAs vary depending on many factors that are significant to microRNA:target interactions. These factors include microRNA:target binding energy, binding location in target sequence, base pairing types between miRNA and target, abundance of miRNAs and targets [@grimson_microrna_2007]. Binding energies of miRNA:target complexes vary based on nucleotide context and determine folding stability of complex [@cao_predicting_2012]. It has been demonstrated that the binding energy between miRNA and target indicates stability or affinity of complex [@helwak_mapping_2013] and does not directly determine repressive activity of miRNA [@cao_predicting_2012]. Early studies have argued that 2-8 nt sequence, seed, located in miRNA 5’UTR bind to specific sequence located in 3’UTR of its target [@bartel_micrornas_2004; @lewis_conserved_2005]. In recent studies, it has been shown miRNAs can interact with targets via sequence that located in different region such as 5’UTR or CDS [@hausser_analysis_2013; @helwak_mapping_2013]moore_mirnatarget_2015. These studies also shown that binding location could indicate functionality of miRNA:target interaction or affect profile of target abundance. It has been shown that miRNAs exhibit repressive activity via, 6-8 nt long sequence that is prefectly complementary with targets, seed region at the 5’UTR of miRNAs [@bartel_micrornas:_2009; @grimson_microrna_2007]. On the other hand, some researchers have reported that seed sequence of miRNA can have mismatches or bulged/wobble nucleotides and locates in different region of miRNAs [@hafner_transcriptome-wide_2010; @helwak_mapping_2013]. In addition to all of these, abundance of miRNAs and targets and miRNA:target ratio in cells are determinant for efficiency of miRNA:target interaction [@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. 
As it is possible for miRNAs to suppress multiple targets, an individual RNA molecule can also be targeted by various miRNAs. In that case, the targeted RNAs exhibit competitor behavior, that is hypothesized as competing endogenous RNAs (ceRNAs) [@ala_integrated_2013; @cesana_deciphering_2013], against their miRNAs. Briefly, Ala et al. have explained the ceRNA hypothesis as disturbtion of the other target when a steady-state system that included one miRNA and two target, was perturbed with expression changes at one of the targets [@ala_integrated_2013]. When interaction between miRNAs and their target were dealed in the cell, explanation and prediction of results of perturbation is difficult due to varieties of interactions. Various computational and experimental studies have applied to be understood for ceRNA:miRNA interactions. It has been observed that when abundance of one of the targets of miR-122 was increased, the other target expression also slightly increased as a result of decreasing repressive activity of miR-122 on them [@denzler_assessing_2014]. Bosson et al. have developed a mathematical model for changes on total target pool concentration after grouping to affinity of targets. They have demonstrated that miRNA activity correlated with affinity between miRNA and target [@bosson_endogenous_2014]. Cooperative efficiency of miRNAs as well as competitor behaviors of targets also were studied and it has been demonstrated crucial for regulations at expression of targets [@denzler_impact_2016]. MiRNA:target interactions have modelled as stoichiometric and catalytic mechanisms and Figliuzzi et al. have recommended to handling models in network structures [@figliuzzi_micrornas_2013]. The model that can explain miRNA target interaction through topological features has been applied at bipartite network by Nitzan et al. [@nitzan_interactions_2014]. Robinson and Henderson applied the model that handles miRNA:target direct and indirect interactions via common miRNA of gene and target of miRNA, at bipartite network. It has been demonstrated that whole miRNAs and targets in network can interacted through common miRNAs and genes [@robinson_modelling_2018]. Associated genes that are targets of the same miRNAs have been found with help of correlation of gene expression changes in recent study [@markus_list_sponge_2017]. List et al. have specified that their approach can be useful for ceRNA’s studies and published these approach as a package that can use with R programming.

@@

We developed a novel network based approach that can use to simulation of gene regulation after any perturbation at miRNA:target interactions. If a perturbation or change induces on graph, whole interacted nodes vary because of changes in miRNA:target ratios. The system was assumed as steady state that miRNAs affect according to the ratio of the individual target to all targets (\autoref{fig1}a, expression values shown as integer for comprehensible sampling). When expression of gene 2 (G2) increased to two fold (\autoref{fig1}b), distribution of miRNA (M1) expression changes (\autoref{fig1}c). Consequently, expression of common target (G4) of miRNA1 and miRNA2 also varies. The change at expression of common target triggers to changes of proportional distribution of miRNA2 (\autoref{fig1}d).

![Shematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangle grey) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed. ](Fig1.jpg){#fig1}

@@

\begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}

\begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

The efficiency of a miRNA on the individual target ($Eff_{gi}$) is calculated according to equation \eqref{eq:1}; $C_m$, miRNA expression in the group; $C_{gi}$, symbolize individual gene expression; $C_g$ gene expressions in the group (Groups are determined depending on miRNAs. Targets that are interacted with the same miRNA find in a group.). In this approach, because a gene can be found in different groups, the cooperative activity of microRNAs also can be calculated. Total repression activity of different miRNAs on the common target, $R$, is calculated based on equation \eqref{eq:2}; sum of efficiency of different miRNAs on the target.

![Calculations for determine to miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE’$, normalized values of seed type efficiency coefficient; $RE’$, normalized values of region efficiency coefficient; $E’$, normalized values of energy coefficient.](fig2.jpg){#fig2}

@@

Interactions between miRNAs and their targets can be affected from different parameters. For this reason, we improved model with factors that can affect to miRNA activity. In our approach, the parameters are classified as binding and efficiency factors and all of these are evaluated in their groups. Binding factors determine interaction between miRNA and target. The efficiency factors decide functionality of binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017]in miRNA:target interactions are represented as binding affinity. Efficiency factors determine how many of microRNA:target complexes will result in inhibition. For example, binding region on the target shows whether the target can be inhibited (functionality)[@hausser_analysis_2013; @helwak_mapping_2013]. Both of binding and efficiency factors are evaluated in the groups and normalized according to equation \eqref{eq:3}; each factor is normalized based on maximum variables that found in their groups.

\begin{equation}
F`= F/F_{max} \tag{3}\label{eq:3}
\end{equation}

The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}).  Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c).

\begin{equation}
Eff_{gi}= C_m * E`_{gi}*STE`_{gi}*C_{gi}/(\sum_{0}^{i} E`_{gi}*STE`_{gi}*C_{gi}) \tag{4}\label{eq:4}
\end{equation}

\begin{equation}
Eff_{gi}= Eff_{gi}*RE`_{gi} \tag{5}\label{eq:5}
\end{equation}

After miRNA binds to its target, but might not repress to bound target. The functionality of bound miRNA on target depends on efficiency factors like region that is binding sequence of miRNA on its target. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:5} (\autoref{fig2}d); $RE’_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in \autoref{fig2}e.

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow.jpg){#fig3}

@@

We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous sections. The functions and steps of approach are explained as following \autoref{fig3}:
1-2. Convertion of dataset: priming_graph function process the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (i-v) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state. 
2-3. Transition of variables in graph: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.
3-4. Trigger change in graph: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The fist, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.
4-5. Updating current values of variables: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.
5-6. Simulation of competing behavior of targets: After the change in the steady- state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.
The node elements in the dataset are handled as two type; repressives (miRNAs) and competings (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the approach are developed with R programming so as can be used with other packages. These are can be found in the github repository [ceRNAnetsim github page](https://github.com/selcenari/ceRNAnetsim) and improved with contributions of others.

@@

We developed a model approach and workflow for competitive ceRNA regulation in this study. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. For this reason, we performed first application as considering only abundance of miRNAs and targets (\autoref{fig1}). After the increasing in a gene expression value, expression values of other genes also changed slightly in comparison with alteration in perturbation started gene node. Different studies have shown that the other gene expression values also rise differently, if a gene abundance increases in ceRNA system [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (\autoref{fig1}d). The common target in system displayed like a trigger for the other group and induced changes of expression values of genes on the other group. Therefore, effect of primary change of the gene (G2) on genes of the other group was observed slightly (\autoref{fig1}d). In addition, like shown by ceRNA hypothesis model of Ala et al., after the increasing of a gene expression (G2), the miRNA that is found in the same group (M1) tended to be more repressive on this target. This caused to decrease in increased expression value of primer triggering gene.
When the expression of a gene (G2) was increased to two fold, we observed that the expression changes were more evident. So, we considered that high miRNA:target ratio results in stronger miRNA repression activity. This results also are coherent with previous reports that were offered by Arvey et al. and Denzler et al.[@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. When the simulation was ran in sample dataset that includes lower target abundance, the first response of primary neihgborhood of the same group was determined as similar with in system that have high target abudance. But regulations of ceRNAs were observed differences in the folowing steps of simulation. In overall state in the end of simulation, we established that all regulations more prominent when compared with low miRNA:target ratio datset. The model appeared the importance of mRNA:target ratio in regulation of competing RNAs.

![Target regulations with interaction parameters. a) In the steady-state the repression activity of miRNAs on the targets after binding and repression efficiency. b) The changes the repression activities after increasing of G2 expression. c) Perturbation of primary neighborhoods of M1 miRNA (M1 miRNA group). d) Regulation of gene expression of other gene group via triggering target (common target between M1 and M2). ](Fig4.jpg){#fig4}

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig4}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig4}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig4} b,c,d).
In regulation of miRNA target sample system (\autoref{fig4}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2.  When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions were observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation.

@@

## References
